---
title: "H3K4me1 vs H3K27ac"
author: "Chenghao Zhu"
date: "1/6/2020"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(reshape2)
library(ggplot2)
library(rtracklayer)
library(biomaRt)
library(ggplot2)
theme_set(theme_classic())
```

Read narrowPeaks

```{r}
h3k4me1_tx = import(
    gzfile("data/H3K4me1_SR8278.merged.nodup_x_ctl_for_rep1.pval0.01.500K.narrowPeak.gz"),
    format = "narrowPeak"
)
h3k4me1_tx = h3k4me1_tx[h3k4me1_tx$qValue > 2]
h3k27ac_tx = import(
    gzfile("data/H3K27ac_SR8278.merged.nodup_x_ctl_for_rep1.pval0.01.500K.narrowPeak.gz"),
    format = "narrowPeak"
)
h3k27ac_tx = h3k27ac_tx[h3k27ac_tx$qValue > 2]
h3k4me1_ve = import(
    gzfile("data/H3K4me1_Vehicle.merged.nodup_x_ctl_for_rep2.pval0.01.500K.narrowPeak.gz"),
    format = "narrowPeak"
)
h3k4me1_ve = h3k4me1_ve[h3k4me1_ve$qValue > 2]
h3k27ac_ve = import(
    gzfile("data/H3K27ac_Vehicle.merged.nodup_x_ctl_for_rep2.pval0.01.500K.narrowPeak.gz"),
    format = "narrowPeak"
)
h3k27ac_ve = h3k27ac_ve[h3k27ac_ve$qValue > 2]
```

## Find overlap peaks

```{r}
overlaps_tx = findOverlapPairs(h3k27ac_tx, h3k4me1_tx)
overlaps_tx
```

There are 12,963 overlapped peaks between h3k27ac and h3k4me1 after treatment. 

```{r}
overlaps_ve = findOverlapPairs(h3k27ac_ve, h3k27ac_ve)
overlaps_ve
```

There are 82,684 overlapped peaks between h3k27ac and h3k4me1 in control. 

```{r}
h3k27ac_tx_overlap = length(unique(overlaps_tx@first$name))
h3k4me1_tx_overlap = length(unique(overlaps_tx@second$name))
h3k27ac_ve_overlap = length(unique(overlaps_ve@first$name))
h3k4me1_ve_overlap = length(unique(overlaps_ve@second$name))
data.frame(
    modification = c("h3k27ac", "h3k27ac", "h3k4me1", "h3k4me1", 
                "h3k27ac", "h3k27ac", "h3k4me1", "h3k4me1"),
    type = c("unique", "overlap", "unique", "overlap",
             "unique", "overlap", "unique", "overlap"),
    treatment = rep(c("tx", "ve"), each = 4),
    npeaks = c(length(h3k27ac_tx) - h3k27ac_tx_overlap, h3k27ac_tx_overlap, 
               length(h3k4me1_tx) - h3k4me1_tx_overlap, h3k4me1_tx_overlap,
               length(h3k27ac_ve) - h3k27ac_ve_overlap, h3k27ac_ve_overlap, 
               length(h3k4me1_ve) - h3k4me1_ve_overlap, h3k4me1_ve_overlap),
    anno.y = c((length(h3k27ac_tx) - h3k27ac_tx_overlap)/2 + h3k27ac_tx_overlap, 
               h3k27ac_tx_overlap/2, 
               (length(h3k4me1_tx) - h3k4me1_tx_overlap)/2 + h3k4me1_tx_overlap, 
               h3k4me1_tx_overlap/2,
               (length(h3k27ac_ve) - h3k27ac_ve_overlap)/2 + h3k27ac_ve_overlap, 
               h3k27ac_ve_overlap/2, 
               (length(h3k4me1_ve) - h3k4me1_ve_overlap)/2 + h3k4me1_ve_overlap, 
               h3k4me1_ve_overlap/2)
) %>%
    mutate(type = relevel(type, ref = "unique"),
           anno.y = anno.y + 100) %>%
    ggplot(aes(x = modification, y = npeaks, fill = type)) +
    geom_col(width = 0.5) +
    geom_text(aes(y = anno.y, label = npeaks)) +
    facet_grid(cols = vars(treatment))
```

For SR8278 treatment, h3k27ac has in total of 84,419 peaks, of which 11,571 overlaps with at least one peak from h3k4m21. h3k4me1 has in total of 93,660 peak, of which 12,353 overlaps with at least one peak from h3k27ac. As for vehicel control, h3k27ac has 82,684 peaks and all of them overlap with a peak from h3k4me1.

The total number of peaks in h3k27ac overlap with h3k4me1 does not match with that of h3k4me1 with h3k27ac, because some peaks overlaps with more than one peak from the other modification.

```{r}
df = merge(
    as.data.frame(table(overlaps_tx@first@seqnames@values)),
    as.data.frame(table(overlaps_ve@first@seqnames@values)),
    by = "Var1"
)
colnames(df) = c("chrom", "tx", "ve")
df %>% filter(chrom %in% paste0("chr", c(1:22, "X", "Y"))) %>% 
    arrange(ve) %>% 
    mutate(chrom = factor(chrom, levels = chrom)) %>%
    melt(id.vars = "chrom", value.name = "freq") %>%
    ggplot() + geom_col(aes(x = chrom, y = freq, fill = variable), position = "dodge") +
    coord_flip()
```

```{r, eval=FALSE, include=FALSE, echo=FALSE}
overlap_size = sapply(seq_along(overlaps_ve), function(i){
    start1 = start(overlaps_ve@first[i])
    end1 = end(overlaps_ve@first[i])
    start2 = start(overlaps_ve@second[i])
    end2 = end(overlaps_ve@second[i])
    if(start1 < start2 & end1 > end2){
        return(end2 - start2)
    } else if(start1 > start2 & end1 < end2){
        return(end1 - start1)
    } else {
        return(min(end1 - start2, end2 - start1))
    }
})

overlap_df = data.frame(
    size = overlap_size,
    peak_dist = abs(start(overlaps_ve@first) + overlaps_ve@first$peak - (start(overlaps_ve@second) + overlaps_ve@second$peak))
)
hist(overlap_size, breaks = 25)
barplot(head(sort(table(overlap_size), decreasing = TRUE), 10))
```

## Gene annotation

Obtain gene annotations

```{r}
chroms_ensembl = seqlevels(h3k27ac_tx)
chroms = gsub("^chr", "", chroms_ensembl)
chroms = gsub("^.+?_", "", chroms)
chroms = gsub("_.+$", "", chroms)
chroms = gsub("v([0-9])$", "\\.\\1", chroms)

ensembl = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
anno = getBM(
    mart = ensembl,
    filter = "chromosome_name",
    values = chroms,
    attributes = c("chromosome_name", "start_position", "end_position", 
                   "ensembl_gene_id", "external_gene_name", "gene_biotype")
)

anno_granges = GRanges(
    seqnames = chroms_ensembl[match(anno$chromosome_name, chroms)],
    ranges = IRanges(
        start = anno$start_position,
        end = anno$end_position,
        names = anno$ensembl_gene_id
    ),
    gene_name = anno$external_gene_name,
    gene_type = anno$gene_biotype
)
```

```{r}
tx_ranges = lapply(seq_along(overlaps_tx), function(i){
    start1 = start(overlaps_tx@first[i])
    end1 = end(overlaps_tx@first[i])
    start2 = start(overlaps_tx@second[i])
    end2 = end(overlaps_tx@second[i])
    res = c(min(start1, start2),
            max(end1, end2))
    names(res) = c("start", "end")
    return(res)
}) %>%
    do.call(rbind, .)
tx_granges = GRanges(seqnames = overlaps_tx@first@seqnames, 
                     ranges = IRanges(start = tx_ranges[,1], end = tx_ranges[,2]))
anno_tx_overlaps = findOverlaps(tx_granges, anno_granges)
anno_tx = anno_granges
anno_tx$n_overlap = as.integer(table(anno_tx_overlaps@to)[as.character(1:length(anno_tx))])
anno_tx = anno_tx[which(!is.na(anno_tx$n_overlap))]
anno_tx = anno_tx[order(anno_tx$n_overlap, decreasing = TRUE)]
anno_tx
```

```{r}
ve_ranges = lapply(seq_along(overlaps_ve), function(i){
    start1 = start(overlaps_ve@first[i])
    end1 = end(overlaps_ve@first[i])
    start2 = start(overlaps_ve@second[i])
    end2 = end(overlaps_ve@second[i])
    res = c(min(start1, start2),
            max(end1, end2))
    names(res) = c("start", "end")
    return(res)
}) %>%
    do.call(rbind, .)
ve_granges = GRanges(seqnames = overlaps_ve@first@seqnames, 
                     ranges = IRanges(start = ve_ranges[,1], end = ve_ranges[,2]))
anno_ve_overlaps = findOverlaps(ve_granges, anno_granges)
anno_ve = anno_granges
anno_ve$n_overlap = as.integer(table(anno_ve_overlaps@to)[as.character(1:length(anno_ve))])
anno_ve = anno_ve[which(!is.na(anno_ve$n_overlap))]
anno_ve = anno_ve[order(anno_ve$n_overlap, decreasing = TRUE)]
anno_ve
```

```{r}
gene_df = left_join(
    data.frame(gene_id   = names(anno_ve),
               gene_name = anno_ve$gene_name, 
               n_overlap = anno_ve$n_overlap,
               stringsAsFactors = FALSE),
    data.frame(gene_id   = names(anno_tx),
               gene_name = anno_tx$gene_name, 
               n_overlap = anno_tx$n_overlap,
               stringsAsFactors = FALSE), 
    by = c("gene_id", "gene_name"), 
)
gene_df$n_overlap.y[is.na(gene_df$n_overlap.y)] = 0
colnames(gene_df) = c("gene_id", "gene_name", "ve", "tx")
gene_df = mutate(gene_df, diff = ve - tx) %>%
    arrange(desc(diff))
gene_df
```

```{r}
gene_df %>% head(20) %>%
    dplyr::select(-diff, -gene_id) %>%
    mutate(gene_name = factor(gene_name, levels = gene_name)) %>%
    melt(id.vars = "gene_name", variable.name = "treatment", value.name = "npeak") %>%
    ggplot() +
    geom_col(aes(gene_name, npeak, fill = treatment), position = "dodge") +
    coord_flip()
```

## GO enrichment

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
gene_down = bitr(gene_df$gene_id[gene_df$diff > 0], fromType = "ENSEMBL", toType = "ENTREZID", OrgDb = org.Hs.eg.db, drop = TRUE)
gene_bg = bitr(gene_df$gene_id, fromType = "ENSEMBL", toType = "ENTREZID", OrgDb = org.Hs.eg.db, drop = TRUE)
ego = enrichGO(
    gene = gene_down$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "ALL",
    pvalueCutoff = 0.01,
    qvalueCutoff = 0.05,
    readable = TRUE
)
ego@result %>% dplyr::select(-ID, -geneID) %>% 
    DT::datatable() %>%
    DT::formatSignif(columns = 5:7, digits = 3)
```




